### 2017-07-07 motivation.tex
EDits to Sec. 2.

To illustrate some of these challenges, consider the behavior of the
\C{new\_order} transaction when executed with a \emph{Read Committed}
(RC) isolation level, the default isolation level in PostgreSQL, a
widely used open-source database system.  An RC transaction is
isolated from \emph{dirty writes}, i.e., writes of uncommitted
transactions, but is allowed to witness the writes of concurrent
transactions as soon as they are committed. Thus, with two concurrent
instances of the \C{new\_order} transaction (call them $T_1$ and
$T_2$), both concurrently placing new orders for different customers
under the same district (\C{d\_id}), PostgreSQL admits the two
executions shown in Fig.~\ref{fig:new_order_execs}.

\begin{figure}[!h]
\centering
\subcaptionbox {
  {\sc rc} Execution 1
  \label{fig:motiv-eg-1-b}
} [
  0.55\columnwidth
] {
  \includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
}
%\hspace*{0.5in}
\subcaptionbox {
  {\sc rc} Execution 2
  \label{fig:motiv-eg-1-a}
}{
  \includegraphics[scale=0.45]{Figures/motiv-eg-1-a}
}
  \caption{\small RC executions involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction depicted in
  Fig.~\ref{fig:new_order_code}. 
  Each instance reads the \C{d\_id} \C{District} record twice, the second
  time to (atomically) update the \C{d\_next\_o\_id} field.}
\label{fig:new_order_execs}
\end{figure}

The figure depicts an execution as a series of transactional read, write, and commit
operations. In the execution on the left, the \C{new\_order} instance
$T_1$ (green) reads the \C{d\_next\_o\_id} field of the district
record for \C{d\_id}, but before it increments the field, another
\C{new\_order} instance ($T_2$) begins its execution and commits. Note
that $T_2$ reads the same \C{d\_next\_o\_id} value as $T_1$, and
inserts new \C{Order} and \C{Order\_line} records with their \C{o\_id}
and \C{ol\_o\_id} fields (resp.) equal to \C{d\_next\_o\_id}. $T_2$
also increments the \C{d\_next\_o\_id} field, which $T_1$ has already
acccessed. This is allowed because reads do not obtain a mutually
exclusive lock on most databases, including PostgreSQL. After $T_2$'s
commit, $T_1$ resumes execution and adds new \C{Order} and
\C{Order\_line} fields with the same order id as $T_1$. Thus, by the
end of the execution, \C{Order\_line} records inserted by $T_1$ and
$T_2$ all bear the same order id. There are also two \C{Order} records
with the same district id (\C{d\_id}) and order id, none of whose
\C{o\_ol\_cnt} reflects the actual number of \C{Order\_line} records
inserted with that order id.  This clearly violates TPC-C's consistency
condition.

Notably, this example does not exhibit any common concurrency bugs
that arise from incorrect use of weak isolation such as
\emph{write-write} conflicts, or \emph{lost updates}.  While $T_1$ and
$T_2$ both increment the \C{d\_next\_o\_id} field of the district
record, they do so atomically (Line 4 of
Fig.~\ref{fig:new_order_code}), allowing both updates to be present in
the final state. Likewise, other well-known anomalies that
characterize RC isolation~\cite{berenson} such as \emph{fuzzy reads},
\emph{phantom reads}, \emph{read skew}, and \emph{write skew}, are
also not exhibited by the example.  Thus, program analyses that aim to
determine appropriate isolation by checking for possible
manifestations of these anomalies would fail to identify grounds for
promoting the isolation level of \C{new\_order} to something stronger.
Yet, if we take the semantics of the application into account, it is
quite clear that RC is not an appropriate isolation level for
\C{new\_order}.

While reasoning in terms of anomalies is cumbersome, and as the above
example shows often inadequate, reasoning about weak isolation in
terms of low-level traces~\cite{adyaphd,gotsmanconcur15} on memory
read and write actions complicates high-level reasoning.  A possible
alternative would interleave weak isolation implementation details
within the program, yielding a (more-or-less) conventional concurrent
program that can be then subject to classical concurrent verification
methods.  Considering the size and complexity of real-world
transaction systems, this strategy is unlikely to scale.

In this paper, we adopt a different approach that \emph{lifts}
isolation semantics (\emph{not} their implementations) to the
application layer, providing a principled framework to simultaneously
reason about application invariants and isolation properties.  To
illustrate this idea informally, consider how we might verify that
\C{new\_order} is sound when executed under a \emph{Repeatable Read}
(RR) isolation level.  PostgreSQL executes an RR transaction by taking
a (conceptual) snapshot of the database state before the transaction
begins that is used by reads performing the transaction.  When the
transaction attempts to update a record, a check is made to determine
if the current version of the record in the database is the same as the
snapshot version. If it is, an exclusive lock is obtained on the
record, and the update is performed. If the record is already locked,
the transaction blocks until the lock is released.  When the lock is
obtained, is held exclusively until the transaction commits or rolls
back. If the aforementioned version check fails (i.e., the database
has a later version than the snapshot), the transaction is rolled back
and re-executed.

\begin{figure}[t]
\includegraphics[scale=0.5]{Figures/RR-postgres}
\caption{Database state transitions corresponding to an execution of
  an RR transaction $T_1$ on PostgreSQL. $T_2$ and $T_3$ are concurrent
  transactions. The database state has three data items, $a$, $b$ and $c$. }
\label{fig:rr-postgres}
\end{figure}
\begin{figure}[t]
\includegraphics[scale=0.5]{Figures/RR-abstract}
  \caption{An abstract execution that includes the concrete execution
  shown in Fig.~\ref{fig:rr-postgres}. It has no locks or snapshots,
  admits fewer interleavings, yet results in the same post-state.  }
\label{fig:rr-abstract}
\end{figure}

Although this implementation, comprising many thousands of lines of
code, is highly involved, its semantic behavior in terms of how it
effects transitions on the database state is fairly simple.  Since a
transaction's reads are always served from a snapshot, no state
changes are witnessed while the execution is in progress. Thus,
insofar as an RR transaction is concerned, the database state does not
change during the execution.  Uncommitted writes are recorded in a
transaction-local state.  When the transaction commits, the local
state is atomically written to the global state to yield a global
state that reflects the transaction's updates.  However, unlike a
strongly isolated serializable transaction, the commit operation is
performed against the current state of the database, not the
snapshot. Thus, after the transaction finishes execution, but before
it commits, the transaction is able to witness the effects of all
concurrent transactions that committed before it.  The PostgreSQL RR
implementation effectively constrains this transition.

We can axiomatize this operational description by observing that (a)
due to the version check, the current transaction cannot update a data
item that was already updated by a concurrent transaction, and (b) due
to the use of exclusive write locks, a data item updated by the
current transaction cannot be overwritten by a concurrent transaction.
If $\Delta$ is the state of the database when an RR transaction
finishes, and $\Delta_c$ is the state visible to the transaction at
the point of commit, we know the transition from $\Delta$ to
$\Delta_c$ (written $\Delta \longrightarrow \Delta_c$) cannot exhibit
effects from any concurrent transactions that write to the same data
items as the current RR transaction.  Similarly, if $\delta$ denotes
a local log that relates transaction variables being written with their updated values,
then $\forall x\in\mathit{dom}(\delta)$, $\Delta_c(x) =
\Delta(x)$. To summarize, the operational semantics of PostgreSQL's RR
implementation can be captured as an axiomatization over transitions of
the database state ($\Delta \longrightarrow \Delta'$) during the
lifetime an RR transaction ($T$):
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}

This simple characterization of RR isolation allows us to verify the
consistency conditions associated with the \C{new\_order}
transaction. First, since the database doesn't change ($\Delta' =
\Delta$) during the execution of the transaction's body, we can reason
about \C{new\_order} as though it executed in complete isolation until
its commit point, leading to a verification process similar to what
would been applied when reasoning about serializability.  When
\C{new\_order} finishes execution and becomes ready to commit, a
transition transfers the transaction's local writes ($\delta$) to the
unchanged database state ($\Delta$).  However, we are required to
consider the interference from concurrent transitions at this point,
which might change the database state from $\Delta$ to $\Delta_c$. If
this interference includes the effects of a concurrent \C{new\_order}
transaction (with the same \C{d\_id}), then verification fails as
described previously.  Fortunately, sequential reasoning shows that
this is impossible - RR prevents a concurrent \C{new\_order}
transaction that modifies the same \C{District} record as the current
transaction (concretely, since the record is already present in the
current transaction's local log, any transition from $\Delta$ to
$\Delta_c$ cannot change this record).  Applying such axiomatic
reasoning on \C{new\_order} allows us to prove that the TPC-C
invariant holds when the transaction is executed under PostgreSQL's RR
isolation.  Our proof framework generalizes this style of reasoning to
various isolation levels on databases.

>>>>

To illustrate some of these challenges, consider the behavior of the
\C{new\_order} transaction when executed with a \emph{Read Committed}
(RC) isolation level, the default isolation level in 8 of the 18
databases studied in~\ref{bailishotos}.  An RC transaction is
isolated from \emph{dirty writes}, i.e., writes of uncommitted
transactions, but is allowed to witness the writes of concurrent
transactions as soon as they are committed. Thus, with two concurrent
instances of the \C{new\_order} transaction (call them $T_1$ and
$T_2$), both concurrently placing new orders for different customers
under the same district (\C{d\_id}), RC isolation allows the
execution shown in Fig.~\ref{fig:new_order_execs}.

\begin{figure}[!t]
\includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
\caption{\small An RC execution involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction depicted in
  Fig.~\ref{fig:new_order_code}. 
  Both instances read the \C{d\_id} \C{District} record concurrently,
  because neither transaction is committed when the reads are
  executed.  The subsequent operations are effectively sequentialized,
  since $T_2$ commits. Nonetheless, reading the same value for
  \C{d\_next\_o\_id} prompts both instances to add \C{Order} records
  with same ids, which inturn triggers the the violation of TPC-C's
  consistency condition.}
\label{fig:new_order_exec}
\end{figure}

% \begin{figure}[!h]
% \centering
% \subcaptionbox {
%   {\sc rc} Execution 1
%   \label{fig:motiv-eg-1-b}
% } [
%   0.55\columnwidth
% ] {
%   \includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
% }
% %\hspace*{0.5in}
% \subcaptionbox {
%   {\sc rc} Execution 2
%   \label{fig:motiv-eg-1-a}
% }{
%   \includegraphics[scale=0.45]{Figures/motiv-eg-1-a}
% }
%   \caption{\small RC executions involving two instances ($T_1$ and
%   $T_2$) of the \C{new\_order} transaction depicted in
%   Fig.~\ref{fig:new_order_code}. 
%   Each instance reads the \C{d\_id} \C{District} record twice, the second
%   time to (atomically) update the \C{d\_next\_o\_id} field.}
% \label{fig:new_order_execs}
% \end{figure}

The figure depicts an execution as a series of SQL 
operations. In the execution, the \C{new\_order} instance
$T_1$ (green) reads the \C{d\_next\_o\_id} field of the district
record for \C{d\_id}, but before it increments the field, another
\C{new\_order} instance ($T_2$) begins its execution and commits. Note
that $T_2$ reads the same \C{d\_next\_o\_id} value as $T_1$, and
inserts new \C{Order} and \C{Order\_line} records with their \C{o\_id}
and \C{ol\_o\_id} fields (resp.) equal to \C{d\_next\_o\_id}. $T_2$
also increments the \C{d\_next\_o\_id} field, which $T_1$ has already
acccessed. This is allowed because reads do not obtain a mutually
exclusive lock on most databases. After $T_2$'s commit, $T_1$ resumes
execution and adds new \C{Order} and \C{Order\_line} fields with the
same order id as $T_1$. Thus, by the end of the execution,
\C{Order\_line} records inserted by $T_1$ and $T_2$ all bear the same
order id. There are also two \C{Order} records with the same district
id (\C{d\_id}) and order id, none of whose \C{o\_ol\_cnt} reflects the
actual number of \C{Order\_line} records inserted with that order id.
This clearly violates TPC-C's consistency condition.

Notably, this example does not exhibit any of the anomalies that
characterize RC isolation~\cite{berenson}. For instance, there are no
\emph{lost writes}; both concurrent transactions' writes are present
in the final state of the database. Thus, program analyses that aim to
determine appropriate isolation by checking for possible
manifestations of these anomalies would fail to identify grounds for
promoting the isolation level of \C{new\_order} to something stronger.
Yet, if we take the semantics of the application into account, it is
quite clear that RC is not an appropriate isolation level for
\C{new\_order}.

While reasoning in terms of anomalies is cumbersome and inadequate,
reasoning about weak isolation in terms of low-level
traces~\cite{adyaphd,gotsmanconcur15} on memory read and write actions
complicates high-level reasoning.  A possible alternative would
interleave weak isolation implementation details within the program,
yielding a (more-or-less) conventional concurrent program that can be
then subject to classical concurrent verification methods.
Considering the size and complexity of real-world transaction systems,
this strategy is unlikely to scale.

In this paper, we adopt a different approach that \emph{lifts}
isolation semantics (\emph{not} their implementations) to the
application layer, providing a principled framework to simultaneously
reason about application invariants and isolation properties.  To
illustrate this idea informally, consider how we might verify that
\C{new\_order} is sound when executed under \emph{Snapshot Isolation}
(SI), and isolation level stronger than RC. Snapshot isolation
allows transactions to be executed against their respective private
snapshots of the database, thus admitting concurrency, but it also
requires that there not be any write-writeconflicts among
concurrent transactions. Write-write conflicts can be eliminated in various
ways; either through detection followed by a rollback, or through
exclusive locks, or a combination of both, leading to different ways
in which SI can be realized. For instance, one possible implementation
of SI, close to the one use by PostgreSQL~\cite{postgres-ssi},
executes a transaction against its private snapshot of the database,
but obtains exclusive locks on the actual records in the database
before performing writes. A write is performed only of the record
hasn't already been updated by a concurrent transaction, i.e,  when it
doesn't result in a write-write conflict.  Otherwise, the transaction is
rolledback. This implementation is demonstrated for a simple database
with three records, $a$, $b$, and $c$ in Fig.~\ref{fig:RR-postgres}.

\begin{figure}[t]
\includegraphics[scale=0.5]{Figures/RR-postgres}
\caption{Database state transitions corresponding to an execution of
  an SI transaction $T_1$ on a PostgreSQL-like store. $T_2$ and $T_3$
  are concurrent transactions. }
\label{fig:rr-postgres}
\end{figure}
\begin{figure}[t]
\includegraphics[scale=0.5]{Figures/RR-abstract}
  \caption{An abstract execution that includes the concrete execution
  shown in Fig.~\ref{fig:rr-postgres}. It has no locks or snapshots,
  admits fewer interleavings, yet results in the same post-state.  }
\label{fig:rr-abstract}
\end{figure}

Implementations of SI on real databases (e.g., PostgreSQL) are
complicated, often running into thousands of lines of code.
Nonetheless, the semantics of Snapshot Isolation, in terms of how it
effects the transitions on the database state, can be captured in a
fairly simple model. Firstly, although SI admits concurrent
transactions, their effects are  not witnessed in the current
transaction (call it $T$), during $T$'s execution. Thus, insofar as
$T$ is concerned, as long as it executes, the global state doesn't
change. More formally, for every consecutive pair of global states
$(\stg,\stg')$ witnessed by $T$ during its execution, $\stg'=\stg$.
After $T$ finishes execution, it commits to the actual database, which
incorporates the effects of concurrent transactions. In executions
where $T$ successfully commits, concurrent transactions are guaranteed
to not be in write-write conflict with $T$. Thus, if $\stg$ is the
global state that $T$ witnessed when it finished execution (which
would be same as the snapshot state), and $\stg'$ is the state to
which $T$ commits, then the diff between $\stg$ and $\stg'$ shouldn't
be in a write-write conflict with $T$. To concretize this notion, let
the database state be a map from transaction variables to values, and
let $\stl$ denote a transaction-local log that maps the variables
being written to their updated values. Then the absence of write-write
conflicts between $T$ and the diff between $\stg$ and $\stg'$ can be
defined as: $\forall x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.
To summarize, the semantics of SI can be captured as an axiomatization
over transitions of the database state ($\Delta \longrightarrow
\Delta'$) during the lifetime an SI transaction ($T$):
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}

% nor $T$'s effects are
% witnessed by concurrent transactions, during $T$'s execution. Thus,
% $T$'s execution can be thought of as a series of transitions of a
% state local to $T$ (call it $\stl$), that has no bearing outside $T$. Conversely, any
% transitions performed by the concurrent transactions on the global
% state (call it $\stg$) has no
% bearing on $T$ since $T$ reads off a static snapshot established at
% the beginning of the transaction. Consequently, any
% state transitions performed by the concurrent transactions can be
% \emph{moved past} $T$'s local state transitions. $T$'s commit,
% however, is a global state transition that is enabled only if
% there are no WW conflicts between $T$ and concurrent transactions.
% Thus, concurrent transactions' state transitions cannot be moved past
% $T$'s commit, and $T$ can be thought of witnessing the effects of
% concurrent transactions just before its commit. 
% therefore treat Since a
% transaction's reads are always served from a snapshot, no state
% changes are witnessed while the execution is in progress. Thus,
% insofar as an RR transaction is concerned, the database state does not
% change during the execution.  Uncommitted writes are recorded in a
% transaction-local state.  When the transaction commits, the local
% state is atomically written to the global state to yield a global
% state that reflects the transaction's updates.  However, unlike a
% strongly isolated serializable transaction, the commit operation is
% performed against the current state of the database, not the
% snapshot. Thus, after the transaction finishes execution, but before
% it commits, the transaction is able to witness the effects of all
% concurrent transactions that committed before it.  The PostgreSQL RR
% implementation effectively constrains this transition.

% We can axiomatize this operational description by observing that (a)
% due to the version check, the current transaction cannot update a data
% item that was already updated by a concurrent transaction, and (b) due
% to the use of exclusive write locks, a data item updated by the
% current transaction cannot be overwritten by a concurrent transaction.
% If $\Delta$ is the state of the database when an RR transaction
% finishes, and $\Delta_c$ is the state visible to the transaction at
% the point of commit, we know the transition from $\Delta$ to
% $\Delta_c$ (written $\Delta \longrightarrow \Delta_c$) cannot exhibit
% effects from any concurrent transactions that write to the same data
% items as the current RR transaction.  Similarly, if $\delta$ denotes
% a local log that relates transaction variables being written with their updated values,
% then $\forall x\in\mathit{dom}(\delta)$, $\Delta_c(x) =
% \Delta(x)$. To summarize, the operational semantics of PostgreSQL's RR
% implementation can be captured as an axiomatization over transitions of
% the database state ($\Delta \longrightarrow \Delta'$) during the
% lifetime an RR transaction ($T$):
% \begin{itemize}
%   \item While $T$ executes, $\Delta' = \Delta$.
%   \item After $T$ finishes execution, but before it commits its local
%     state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
% \end{itemize}

This simple characterization of SI isolation allows us to verify the
consistency conditions associated with the \C{new\_order} transaction.
First, since the database doesn't change ($\Delta' = \Delta$) during
the execution of the transaction's body, we can reason about
\C{new\_order} as though it executed in complete isolation until its
commit point, leading to a verification process similar to what would
been applied when reasoning about serializability.  When
\C{new\_order} finishes execution, but before it commits, the SI
axiomatization shown above requires us to consider global state
transitions $\stg \longrightarrow \stg'$ that do not include changes
to the records ($\stl$) written by \C{new\_order}, i.e.,
$\forall(x\in\delta).~\Delta'(x) = \Delta(x)$. Since concurrent
\C{new\_order} transactions that write to the same C{District} record
modify the record by incrementing its \C{d\_next\_o\_id} field,
their transitions do not satisfy the necessary condition imposed by
the SI axiomatization, allowing us to rightfully ignore their
interference. This leaves the interference due to concurrent
\C{new\_order} transactions that write to a different district record. 
Fortunately, such interference does not prompt \C{new\_order} to
violate the invariant, which can be established by reasoning on the
similar lines as concurrent program logics, such as the Rely-Guarantee
logic.

% and becomes ready to commit, a
% transition transfers the transaction's local writes ($\delta$) to the
% unchanged database state ($\Delta$).  However, we are required to
% consider the interference from concurrent transitions at this point,
% which might change the database state from $\Delta$ to $\Delta_c$. If
% this interference includes the effects of a concurrent \C{new\_order}
% transaction (with the same \C{d\_id}), then verification fails as
% described previously.  Fortunately, sequential reasoning shows that
% this is impossible - RR prevents a concurrent \C{new\_order}
% transaction that modifies the same \C{District} record as the current
% transaction (concretely, since the record is already present in the
% current transaction's local log, any transition from $\Delta$ to
% $\Delta_c$ cannot change this record).  Applying such axiomatic
% reasoning on \C{new\_order} allows us to prove that the TPC-C
% invariant holds when the transaction is executed under PostgreSQL's RR
% isolation.  Our proof framework generalizes this style of reasoning to
% various isolation levels on databases.
